---
title: "Statistical Modelling Assignment 3"
author: "Karl Evans"
date: "18/05/2022"
output: 
  pdf_document:
    latex_engine: xelatex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, tidymodels, gglm)
```

# Q1
## a)
i. $\log(\lambda_{11}) = \gamma_0$  
ii. $\log(\lambda_{12}) = \gamma_0+\gamma_2$  
iii. $\log(\lambda_{21}) = \gamma_0+\gamma_1$  
iv. $\log(\lambda_{22}) = \gamma_0+\gamma_1+\gamma_3$  

## b)
$Y_{i2}|Y_{i1}+Y_{i2}=n_i$ is the number of cases per total population in each treatment group.  
A Case is then success from a fixed number of bernoulli trials with fixed probability of success and so has a binomial distribution.  
$\pi_i$ is the probability of cases for each treatment group, taken as the proportion: $\pi_i = \frac{\lambda_{i2}}{\lambda_{i1}+\lambda_{i2}}$

## c)
$\gamma_0$ is the intercept, common to all groups.  
$\gamma_1$ changes the control rate between treatment groups.  
$\gamma_2$ changes the case rate equally for both treatment groups.  
$\gamma_3$ changes the case rate only for treatment B, so creates the difference in cases BETWEEN treatment groups.  
Therefore, $\beta_1$ of the logistic model is only affected by $\gamma_3$ in the poisson model.

## d)
So to test the effect of treatment A vs B:  
 - the probability, $\pi_i$ of a case changes if $\beta_1$ is significant.  
 - the case rates compared to baseline rates within the treatment group $\lambda_{12}$ vs. $\lambda_{12}+\lambda_{11}, \lambda_{22}$ vs. $\lambda_{22}+\lambda_{21}$ if the interaction term $\gamma_3$ is significant.

\newpage

# Q2
## a) Read Data
```{r, message=FALSE}
lung<-read_csv("lung_cancer.csv")
```

## b) EDA
```{r}
lung <- 
  lung %>% 
  mutate(
    across(where(is.character), factor)
  )
lung<-lung %>% 
  mutate(p=cases/pop)
skimr::skim(lung)
```

```{r, fig.cap="Histogram of Cases", fig.height = 2, fig.width = 3, fig.align = "center", message=FALSE}
lung %>% 
  ggplot(aes(cases)) + 
  geom_histogram(col = "black", fill = "orange")
```

```{r, fig.cap="Visualising Cases VS Age and City", fig.height = 11, fig.width = 6, fig.align = "center"}
p1<-lung%>%
  ggplot(aes(age,cases, col=city, shape=city))+
  geom_point(size=4)+
  labs(title="Scatterplot of Cases by Age Group", x = "Age Goup", y = "Number of Cases")+
  scale_shape()
p2<-lung%>%
  ggplot(aes(city,cases, col=age, fill=age))+
  geom_point(position=position_jitter(h=0, w=0.2),
             shape = 21, alpha = 0.5, size = 3)+
  labs(title="Scatterplot of Cases by Age Group", x = "Age Goup", y = "Number of Cases")
p3<-lung%>%
  ggplot(aes(age, cases))+
  geom_boxplot()+
  labs(title="Boxplot of Cases by Age Group", x = "Age Goup", y = "Number of Cases")
p4<-lung%>%
  ggplot(aes(city, cases))+
  geom_boxplot()+
  labs(title="Boxplot of Cases by City", x = "City", y="Number of Cases")
gridExtra::grid.arrange(p1,p2,p3,p4, ncol=1)
```

```{r, fig.cap="Visualising Proportion of Cases Vs Age and City", fig.height = 11, fig.width = 6, fig.align = "center"}
p1<-lung %>%
  ggplot(aes(age,p, col=city))+
  geom_point()+
  labs(title="Scatterplot of Proportion of Cases Vs Age", y="Proportion of Cases",x = "Age Group")
p2<-lung %>%
  ggplot(aes(city,p, col=age))+
  geom_point()+
  labs(title="Scatterplot of Proportion of Cases Vs City", y="Proportion of Cases",x = "City")
p3<-lung%>%
  ggplot(aes(age, p))+
  geom_boxplot()+
  labs(title="Boxplot of Proportion of Cases by Age", y="Proportion of Cases",x = "Age Group")
p4<-lung%>%
  ggplot(aes(city, p))+
  geom_boxplot()+
  labs(title="Boxplot of Proportion of Cases by City", y="Proportion of Cases",x = "City")
gridExtra::grid.arrange(p1,p2,p3,p4, ncol=1)
```

\newpage
## c) Model 1
```{r}
M1<-glm(cases~1, family=poisson, offset=log(pop) ,data=lung)
summary(M1)
```

## d) Model 2
```{r}
M2<-glm(cases~city+age, family=poisson, offset=log(pop) ,data=lung)
summary(M2)
```

## e) Model 3
```{r}
M3<-glm(cases~age+city+log(pop), family=poisson, offset=log(pop) ,data=lung)
summary(M3)
```

\newpage
## f) ANOVA
```{r}
anova(M1,M2, test='LRT')
```
With a p-value << 0.05 on the Likelihood Ratio Test, city and age are significant predictors of the number of cases.  

## g) AIC

```{r}
AIC(M1,M2,M3)
```


## h) Coefficients of M2
```{r}
coef(M2)
```

## i) Pearson Residuals
```{r}
lung<-lung %>% add_column(M2_res = residuals(M2, type = "pearson"), M2_fit = fitted(M2))
```
## j) Residuals Plots
```{r, fig.cap="Residual Plots" , message=FALSE, fig.height = 6, fig.width = 4, fig.align = "center"}
p1<-lung %>%
  ggplot(aes(M2_fit,M2_res))+
  geom_point()+
  geom_smooth()+
 labs(x="Fitted Values", y="Pearson Residuals", title="Pearson Residuals Vs Fitted Values")
p2<-lung %>%
  ggplot(aes(age,M2_res))+
  geom_point()+
 labs(x="Age Group", y="Pearson Residuals", title="Pearson Residuals Vs Age Group")
p3<-lung %>%
  ggplot(aes(city,M2_res))+
  geom_point()+
    geom_smooth()+
 labs(x="City", y="Pearson Residuals", title="Pearson Residuals Vs City")
gridExtra::grid.arrange(p1,p2,p3, ncol=1)
```

```{r, fig.cap="Residual Plots"}
lung %>%
  select(where(is.factor), M2_res) %>% 
  pivot_longer(-M2_res) %>% 
  ggplot(aes(value, M2_res)) + 
  geom_boxplot() + 
  facet_wrap(~name, scale = "free")+
  labs(y="Pearson Residual", x="Predictor")
```

\newpage

## k) Prediction
```{r}
new_data <- 
  tibble(
    city = as_factor("Fredericia"),
    age = as_factor("40-54"),
    pop=4000,
  )
new_data
```

The probability of 5 or less cases is binomial (n=population, $p=\frac{\lambda}{population}$)
```{r}
lambda=exp(predict(M2, newdata = new_data))
rate=lambda/4000
pbinom(5, 4000, rate)
```

The probability of 5 or less cases, given the rate from data modeled by M2, is ~0.0044. Since this is less than p=0.05, it is unlikely that with 5 cases, the rate for Fredericia in 1980 is the same as the modeled data. Therefore it is a significant decrease in the rate of lung cancer cases.

